---
title: "Creating Modular Repoorting with heddlr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating Modular Repoorting with heddlr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This document serves as an introduction to using `heddlr`, a set of tools to 
help dynamically build R Markdown documents based on repeating sections, for 
somewhat more complicated reports. Through the course of this vignette, we'll 
be going over the core functions which make up `heddlr` and how they interact 
to help make report generation faster.

One of the fundamental concepts of this package is that `heddlr` is content 
agnostic -- that is, it should be able to cooperate with any and all R 
(and therefore, R Markdown) extensions, which it just views as strings to 
manipulate. This means that our first step in demonstrating `heddlr` will 
actually require us to write some other code beforehand.

We're going to be using `heddlr` to make a `flexdashboard` using the 
`nycflights13` data set. In order to do that, we'll first need to load a few
packages:

```{r setup}
library(heddlr)
library(dplyr) # For our data manipulation
library(ggplot2) # For graphing utilities
library(nycflights13) # Provides the flight dataset
library(ggalluvial) # Helps build one of our charts
```

You'll also need to have `purrr` and `tidyr` installed to run this document 
completely -- however, we won't load them outright in this document.

`heddlr` is designed to help build documents with a number of repeating 
sections, which would normally require copying the same code over and over
again. Even before using any of the new tools in `heddlr`, we would be able to
reduce this repetition somewhat by extracting the repeated pieces into 
functions. Our dashboard will be built from four graphs, each of which 
will be repeated some number of times -- we'll go ahead now and build out a 
function for each of those graphics.

```{r}
graph_flights_per_hour <- function(origin_code) {
 flights %>%
  filter(origin == origin_code) %>%
  count(time_hour, hour) %>%
  group_by(hour) %>%
  summarise(n = mean(n)) %>%
  ggplot(aes(hour, n, fill = n)) +
  geom_col(color = "black") + 
  scale_x_continuous(breaks = seq(2, 24, 2)) +
  scale_fill_viridis_c(begin = 0, end = 0.7, name = "") +
  theme_minimal() %+replace%
  theme(axis.text.y = element_blank()) +
  coord_polar() + 
  labs(title = "Average number of flights per hour",
       subtitle = paste(origin_code, "Airport"),
       x = "",
       y = "") 
}
graph_flights_per_hour("EWR")
```

```{r}
graph_mean_delay <- function(origin_code, month_number) {
 flights %>%
  filter(origin == origin_code &
           month == month_number) %>%
    mutate(date = lubridate::as_date(time_hour)) %>%
  group_by(date) %>%
  summarise(delay = mean(dep_delay, na.rm = T)) %>%
  ggplot(aes(date, delay)) + 
  geom_col(color = "black") + 
  scale_y_continuous(expand = expand_scale(mult = c(0, 0.07))) + 
  theme_classic() + 
  labs(x = "Day of month",
       y = "Delay (minutes)",
       title = "Average flight departure delays per day",
       subtitle = paste(origin_code, "Airport,", month.name[[month_number]]))
}

graph_mean_delay("EWR", 1)
```

```{r}
graph_sankey <- function(origin_code) {
  flights %>%
  count(origin, dest) %>%
  mutate(highlight = ifelse(origin == origin_code, 1, 0)) %>%
  ggplot(aes(axis1 = origin, axis2 = dest, y = n)) +
  geom_alluvium(aes(fill = highlight)) + 
  geom_stratum(width = 1/10) + 
    geom_label(stat = "stratum", infer.label = TRUE) +
  scale_y_continuous(expand = expand_scale(mult = c(0, 0.07))) + 
  theme_classic() %+replace%
  theme(legend.position = "none",
        axis.text = element_blank()) + 
  labs(x = "",
       y = "",
       title = paste("Flights out of", "EWR"))
}

graph_sankey("EWR")
```

```{r}
graph_carriers <- function(origin_code) {
 flights %>%
  filter(origin == origin_code) %>%
  group_by(carrier) %>%
  summarise(dep_delay = mean(dep_delay, na.rm = T)) %>%
  ggplot(aes(carrier, dep_delay)) +
  geom_col() + 
  scale_y_continuous(expand = expand_scale(mult = c(0, 0.07))) + 
  theme_classic() + 
  labs(x = "Carrier",
       y = "Average departure delay (minutes)",
       title = "Mean departure delay by carrier",
       subtitle = paste("EWR", "Airport")) 
}

graph_carriers("EWR")
```

So now we have the functions that will produce our actual content for the 
dashboard. Our goal is to repeat each of the graphics for each of the three
origin airports in our dataset, and to repeat the monthly delay graph once for
each month at each of these airports.

This is where we'll start using functions from `heddlr`. In `heddlr` 
terminology, each one of those repeated sections is what we'll call a 
`pattern`. We typically store those patterns in their own files -- that makes
it a bit easier to reason about what each one needs to include and how they'll
piece together. For this dashboard, we've saved off two patterns -- one is 
called `flights_site_level.Rmd`, and the other is `flights_month_level.Rmd`.

We have a few ways we can then import those patterns into our R session to 
work with. The first and simplest of these is `import_pattern`, which just 
takes as argument a file path:

```{r}
pattern <- import_pattern("flights_site_level.Rmd")
pattern
```

Of course, we usually have more than one pattern to work with, and it can be
nice to keep them all in one place. `heddlr` helps you do this as well via the
`assemble_draft` function, which takes named filepaths and stores them in a list:

```{r}
draft <- assemble_draft("site_pattern" = "flights_site_level.Rmd",
               "month_pattern" = "flights_month_level.Rmd")
draft
```

There's one last function in `heddlr` to help us create pattern objects, named
`create_yaml_header`. If we wanted, we can include a pattern in our 
`assemble_draft` call that contains the YAML needed to render our R Markdown
document. However, `create_yaml_header` also provides us the option of 
creating that header natively in our generator document by passing it a list:

```{r}
dashboard_header <- list(
  "title" = "2013 NYC Flights Dashboard",
  "author" = "Mike Mahoney",
  params = list(
    "flights" = "NA",
    "graph_flights_per_hour" = "NA",
    "graph_mean_delay" = "NA",
    "graph_sankey" = "NA",
    "graph_carriers" = "NA"
  ),
  "output" = list("flexdashboard::flex_dashboard" = list(
    "vertical_layout" = "fill",
        "orientation" = "rows"
  ))
)

dashboard_header <- create_yaml_header(dashboard_header)
dashboard_header
```

Now that we have our patterns in R, it's time to use them to generate our 
dashboard. In each of our patterns, I've gone ahead and used placeholder values
in both the code and markdown to indicate where I want to use values instead. 
We can then go ahead and use the `heddle` function to replace those values 
appropriately.

The `heddle` function takes three key arguments -- our data, the pattern we're
working with, and the placeholder values to replace. If the data we pass is
a vector (like in a basic `mutate` call), that last argument can just be all 
the placeholders we want to replace -- heddlr will automatically assume you 
want to replace it with the values in the vector. If we pass it a dataframe 
instead, we need to name our arguments in the format `"Placeholder" = Variable`
-- which allows us to use multiple different variables to replace the 
placeholders.

```{r}
dashboard_body <- flights %>%
  distinct(origin, month) %>%
  arrange(month) %>%
  tidyr::nest(month = month) %>%
  mutate(sitepattern = heddle(origin, draft$site_pattern, "AIRCODE"),
         monthpattern = purrr::map(month, heddle, draft$month_pattern, "MONTHNUM" = month))

dashboard_body
```

You'll notice that even though we made both our site and month patterns with 
`heddle`, "sitepattern" came out as a single text string, while "monthpattern"
is currently saved as a list. That brings us to our next function,
`make_template`. This function can be used in a few different ways:

* If used with `mutate` and `purrr::map`, it can be used to flatten those
  list vectors into text strings, which is necessary for exporting our 
  dashboard
* If used after a pipe step, it collapses a template vector into a single
  string
* If used on its own, it can combine several different template objects
  into one
  
We'll use the first two right now to get the majority of our dashboard into
one object, which will form the bulk of our dashboard:

```{r}
dashboard_body <- dashboard_body %>%
  mutate(monthpattern = purrr::map_chr(monthpattern, make_template),
         template = paste0(sitepattern, monthpattern)) %>%
  make_template(template)

substr(dashboard_body, 1, 40)
```

So now we have two main dashboard objects -- the dashboard YAML header we
created earlier, and the dashboard body we just finished. We'll use the last
use of `make_template` to combine the two of those -- note that `make_template`
can't combine separate objects when used in a pipe, due to some limitations
in how it looks for the objects in your R environment.

We're also going to add in the last major `heddlr` function: `export_template`.
By providing a template object and a file path to `export_template`, you'll 
wind up with a .Rmd document ready to be knit into its final output.

```{r}
make_template(dashboard_header, dashboard_body) %>%
  export_template("flights_dashboard.Rmd")
```

We're then able to call `rmarkdown::render` on our new .Rmd document from the
same document we used to generate it. Because we've already gone ahead and 
defined our datasets and functions, we're able to knit the document with 
parameters to speed up processing time -- this is particularly helpful if your
data is particularly large or coming from a SQL connection, so you only need
to import the data once. I included the parameters in our `create_yaml_header`
step, and will include them in this render call as well:

```{r results="hide"}
rmarkdown::render("flights_dashboard.Rmd",
                  params = list(
                    flights = flights,
                    graph_flights_per_hour = graph_flights_per_hour,
  graph_mean_delay = graph_mean_delay,
  graph_sankey = graph_sankey,
  graph_carriers = graph_carriers
                  ))
```

We've now fully automated the building and rendering of our dashboard, using
about 100 lines of code to quickly piece together an R Markdown report which
can be arbitrarily long, with sections defined by our dataset. The hope is that
this helps create reports for large reports or reocurring reporting based on 
shifting data sets.
